---
title: "AnonymousCompany Project - Time Series"
author: "Luke Stucky"
theme: cosmo
format: dashboard
logo: blue_drop.png
editor: visual
---

### Loading in datasets

```{r}
#| include: false

load("AnonymousCompany_DSRATING_APPEND_20240401.RData")
DS <- dsdata2
load("AnonymousCompany_GIVING_CLEAN_USONLY_20240401.RData")
Giving <- gs5
load("state_abbrev.RData")
State <- state_abbrev
```

### Packages

```{r}
#| include: false

library(dplyr)
library(tidyverse)
library(plotly)
library(ggplot2)
library(plotly)
library(RColorBrewer)
library(tidyverse)
library(scales)
```

# Total Donation Trend

## Row {height="20%"}

```{r}
#| include: false
combined_data <- merge(DS, Giving, by = "accountid", all = TRUE)
```

```{r}
#| content: valuebox
#| title: "Lifetime Donations"

list(color = "green",
  value = dollar(sum(Giving$giving_lifetime))
)
```

```{r}
#| content: valuebox
#| title: "Total Donations in 2017"

list(color = "yellow",
  value = dollar(sum(Giving$giving2017))
)
```

```{r}
#| content: valuebox
#| title: "Total Donations in 2021"

list(color = "blue",
  value = dollar(sum(Giving$giving2021))
)
```

## Row {height="80%"}

### Column {width="40%"}

```{r}
#| title: "Total Giving Each Year"

# Total Giving Each Year Function
calculate_total_giving <- function(data) {
  total_giving <- Giving %>%
    summarise(
      giving2017 = sum(giving2017, na.rm = TRUE),
      giving2018 = sum(giving2018, na.rm = TRUE),
      giving2019 = sum(giving2019, na.rm = TRUE),
      giving2020 = sum(giving2020, na.rm = TRUE),
      giving2021 = sum(giving2021, na.rm = TRUE)
    ) %>%
    pivot_longer(cols = starts_with("giving"), names_to = "year", values_to = "total_giving")
  
  return(total_giving)
}

# Total Giving Each Year Calculated
total_giving <- calculate_total_giving(Giving)

# Select Colors
color_palette <- brewer.pal(5, "Set3")

# Create Line Chart
line_chart <- plot_ly(total_giving, x = ~year, y = ~total_giving, type = 'scatter', mode = 'lines+markers',
                      line = list(color = color_palette[1], width = 5)) %>%
  layout(title = "Total Giving Each Year",
         xaxis = list(title = "Year"),
         yaxis = list(title = "Total Giving"))

# Line Chart Output
line_chart
```

**Key Findings:** - Total giving was increasing steadily until 2020, and then it fell.\
- Despite the fall in donations in 2020, total donations in 2021 were able to recover.\

# DS Ratings

## Row {height="20%"}

#### Column {.tabset}

```{r}
#| content: text
#| title: "DS 1-1"
cat("High-quality match as a donor giving a single gift of $5,000 or more to a nonprofit or political organization found in DonorSearch’s charitable giving database. A DS1-1 rated prospect may have markers of wealth or may only be matched to a significant giving history, and have no notable wealth markers.")

```

```{r}
#| content: text
#| title: "DS 1-2"
cat("High-quality match to one or more exceptional markers of wealth including Real Estate holdings of $2+ Million, business executives at a firm with revenues of $5+ Million, a Foundation Trustee, SEC Insider and/or cumulative Federal political giving of $10k+.")

```

```{r}
#| content: text
#| title: "DS 1-3"
cat("High-quality match to one or more lower--but notable--markers of wealth, including Real Estate holdings of $1-2 Million, a business executive at a firm with revenues of $1-5 Million, and/or cumulative Federal political giving of $5-$10k+.")

```

```{r}
#| content: text
#| title: "DS 1-4"
cat("High-quality matches to one or more significant markers of wealth, including Real Estate holdings of $500,000-1 Million, ownership of at least three properties or a real estate trust, and/or a single gift of over $500.")
```

```{r}
#| content: text
#| title: "DS 1-5"
cat("High-quality matches to markers of wealth or philanthropy, including one or more of the following: A business executive at a firm with revenues of $1 Million, or the strong possibility of being an SEC insider and/or member of a foundation board.")
```

```{r}
#| content: text
#| title: "DS 2"
cat("Exact match to selected markers of wealth including Airplane or Boat ownership and/or a holder of a pilot’s license.")
```

```{r}
#| content: text
#| title: "DS 3"
cat("No noteworthy matches to giving history or wealth indicators. Note: If the name or address matched is inaccurate, the rating is automatically DS 3.")
```

## Row {height="80%"}

### Column {width="40%"}

### Column {.tabset}

```{r}
#| title: "Average Giving for DS Ratings"


# DS Ratings to include
desired_ds_ratings <- c("DS1-1", "DS1-2", "DS1-3", "DS1-4", "DS1-5", "DS2", "DS3")

# DS Ratings Factored
combined_data$ds_rating <- factor(combined_data$ds_rating)

# Group By DS Ratings and Find Average Giving of Each Rating
average_giving <- combined_data %>%
  filter(!is.na(ds_rating)) %>%  # Exclude rows with NA in ds_rating
  filter(ds_rating %in% desired_ds_ratings) %>%  # Filter desired DS ratings
  group_by(ds_rating) %>%
  summarise(average_giving = mean(giving_lifetime, na.rm = TRUE),
            count_accounts = n())

# Select Colors
colors <- brewer.pal(12, "Paired")

# Bar Chart Created
p <- ggplot(average_giving, aes(x = ds_rating, y = average_giving, text = paste("<br>Accounts: ", count_accounts))) +
  geom_bar(stat = "identity", fill = colors[3]) +  # Use the first color from Set3 palette
  labs(x = "DS Rating", y = "Average Lifetime Giving Amount", title = "Average Lifetime Giving Amount by DS Rating") +
  theme_minimal()

# Interactive Chart Output
ggplotly(p)
```

**Key Findings:**\
- DS1-1, DS1-2, and DS1-3 have the highest average giving amount, but they have fewer accounts than DS1-4, DS1-5, and DS3.\
- While some ratings may have a higher average giving amount, they may not have as many accounts in their rating.\
- The DS2 rating (no noteworthy matches to giving history/wealth) has the lowest average lifetime giving amount of less than \$1,000.\

### Column {.tabset}

```{r}
#| title: "Average Giving by Year for Top Five DS Ratings"
#| include: false
# Define the desired DS ratings
desired_ds_ratings <- c("DS1-1", "DS1-2", "DS1-3", "DS1-4", "DS1-5", "DS2", "DS3")

# Top Five DS Rating Donors
top_ratings <- combined_data %>%
  filter(!is.na(ds_rating)) %>% 
  group_by(ds_rating) %>%
  summarise(average_giving = mean(giving_lifetime, na.rm = TRUE)) %>%
  filter(ds_rating %in% desired_ds_ratings) %>%  # Filter desired DS ratings
  top_n(7, average_giving) %>%
  arrange(desc(average_giving)) %>%
  pull(ds_rating)

# Top DS Rating Donor Data
top_ratings_data <- combined_data %>%
  filter(ds_rating %in% top_ratings)

# Calculate Average Giving by Year for Top Five DS Ratings
average_giving_by_year <- top_ratings_data %>%
  group_by(ds_rating) %>%
  summarise(
    mean_2017 = mean(giving2017, na.rm = TRUE),
    mean_2018 = mean(giving2018, na.rm = TRUE),
    mean_2019 = mean(giving2019, na.rm = TRUE),
    mean_2020 = mean(giving2020, na.rm = TRUE),
    mean_2021 = mean(giving2021, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("mean"), names_to = "year", values_to = "average_giving")

# Create Line Chart
p <- ggplot(average_giving_by_year, aes(x = year, y = average_giving, color = ds_rating, group = ds_rating)) +
  geom_line(size = 1.5) +
  geom_point() +
  labs(title = "Average Giving by Year for Top Five DS Ratings",
       x = "Year",
       y = "Average Giving",
       color = "DS Rating") +
  scale_color_brewer(palette = "Paired") +
  theme_minimal()

# Create Interactive Chart
ggplotly(p)
```

```{r}
#| title: "Top Donating DS Ratings Each Year"

# Define the desired DS ratings
desired_ds_ratings <- c("DS1-1", "DS1-2", "DS1-3", "DS1-4", "DS1-5", "DS2", "DS3")

# Top Five DS Rating Donors
top_ratings <- combined_data %>%
  filter(!is.na(ds_rating)) %>% 
  group_by(ds_rating) %>%
  summarise(average_giving = mean(giving_lifetime, na.rm = TRUE)) %>%
  filter(ds_rating %in% desired_ds_ratings) %>%  # Filter desired DS ratings
  top_n(7, average_giving) %>%
  arrange(desc(average_giving)) %>%
  pull(ds_rating)

# Top DS Rating Donor Data
top_ratings_data <- combined_data %>%
  filter(ds_rating %in% top_ratings)

# Subset by Year
giving_by_year <- top_ratings_data %>%
  group_by(ds_rating) %>%
  summarise(
    giving2017 = sum(giving2017, na.rm = TRUE),
    giving2018 = sum(giving2018, na.rm = TRUE),
    giving2019 = sum(giving2019, na.rm = TRUE),
    giving2020 = sum(giving2020, na.rm = TRUE),
    giving2021 = sum(giving2021, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = starts_with("giving"), names_to = "year", values_to = "total_giving")

# Create Line Chart
p <- ggplot(giving_by_year, aes(x = year, y = total_giving, color = ds_rating, group = ds_rating)) +
  geom_line(size = 2) +
  geom_point() +
  labs(title = "Total Giving by Year for Top DS Ratings",
       x = "Year",
       y = "Total Giving",
       color = "DS Rating") +
  scale_color_brewer(palette = "Paired") +
  theme_minimal()

# Create Interactive Chart
ggplotly(p)

```

**Key Findings:**\
- DS1-1, DS1-2, DS1-3, and DS3 stay consistent in their total giving each year, but they are not the top total donors.\
- DS1-5 has been the second highest donor each year, and in 2021, they become the highest. It appears that they will stay ahead in upcoming years.\
- The DS1-4 rating has been the highest donor up until 2021, where DS1-5 goes ahead of them.\
- DS2 only has three in their category, so they make up a very small amount of the total giving.

# Account Type

## Row {width="20%"}

```{r}
#| include: false
# Calculate Total Lifetime Giving for Households and Organizations
total_lifetime_giving <- combined_data %>%
  group_by(account_type) %>%
  summarise(lifetime_giving = sum(giving_lifetime, na.rm = TRUE))

# Filter Data for Household and Organization
household_lifetime_giving <- total_lifetime_giving %>%
  filter(account_type == "Household Account") %>%
  pull(lifetime_giving)

organization_lifetime_giving <- total_lifetime_giving %>%
  filter(account_type == "Organization") %>%
  pull(lifetime_giving)
```

```{r}
#| content: valuebox
#| title: "Household Donations"
list(color = "blue",
  value = dollar(sum(household_lifetime_giving))
)

```

```{r}
#| content: valuebox
#| title: "Organization Donations"

list(color = "green",
  value = dollar(sum(organization_lifetime_giving))
)
```

## Row {height="80%"}

### Column {width="40%"}

### Column {.tabset}

```{r}
#| title: "Organizations versus Households Donation Percentages"

# Total Donations Each Year
combined_data <- combined_data %>%
  mutate(total_donations = rowSums(select(., c("giving2017", "giving2018", "giving2019", "giving2020", "giving2021")), na.rm = TRUE))

# Total Donations By Account Type Each Year
account_type_total <- combined_data %>%
  pivot_longer(cols = c(giving2017, giving2018, giving2019, giving2020, giving2021),
               names_to = "year",
               values_to = "total_giving") %>%
  filter(!is.na(account_type)) %>%
  group_by(year, account_type) %>%
  summarise(total_giving = sum(total_giving, na.rm = TRUE)) %>%
  group_by(year) %>%
  mutate(percentage = total_giving / sum(total_giving) * 100) %>%
  ungroup()

# Select Colors
colors <- brewer.pal(12, "Paired")

# Create Dodged Bar Chart
p <- ggplot(account_type_total, aes(x = year, y = percentage, fill = account_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Percentage of Total Donations by Account Type Each Year",
       x = "Year",
       y = "Percentage of Total Donations") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = colors[2:3], labels = c("Household", "Organization"))

# Create Interactive Chart
ggplotly(p)

```

**Key Findings:**\
- Household Accounts consistently contribute the majority of total donations across all years.\
- It appears that households were pulling away and donating a lot more than organizations up until 2021.\
- While households make up the majority of donations currently, it could change in the upcoming years as we are starting to see a shift in the year 2021.\

### Column {.tabset}

```{r}
#| title: Organizations versus Households Consistently

# Create Consistency Variable
combined_data <- combined_data %>%
  mutate(
    total_donations = rowSums(select(., starts_with("giving")) > 0),
    consecutive_donations = rowSums(select(., starts_with("giving")) > 0) == 5,
    consistency_rank = case_when(
      consecutive_donations ~ "Very High",
      total_donations >= 4 | total_donations == 3 ~ "High",
      total_donations == 2 | total_donations == 3 ~ "Medium",
      TRUE ~ "Low"
    )
  )

# Filter out NA values in account_type
combined_data <- combined_data[!is.na(combined_data$account_type),]

# Specify the order of levels for consistency_rank
combined_data$consistency_rank <- factor(combined_data$consistency_rank, levels = c("Low", "Medium", "High", "Very High"))
# Select Colors
colors <- brewer.pal(12, "Paired")

# Create Dodged Bar Chart
chart_dodge <- combined_data %>%
  group_by(account_type, consistency_rank) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  ggplot(aes(x = consistency_rank, y = percentage, fill = factor(account_type))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Comparison of Consistency Rank between Households and Organizations",
       x = "Consistency Rank",
       y = "Percentage of Account IDs",
       fill = "Account Type") +
  theme_minimal() +
  scale_fill_manual(values = colors[2:3], labels = c("Household", "Organization"))

# Create Interactive Chart
ggplotly(chart_dodge)
```

**Key Findings:**\
- Both organizations and households follow a similar trend. They both have a small percentage of rank 1, then their highest consistency rank is 2, and then they decrease.\
- Rank 2 is the most common for accounts to be in.\
- Organizations tend to donate more consistently, as they have a higher percentage of accounts in ranks 3 and 4.\

# State Trends

```{r}
#| title: "Most Valuable States"

# Find Total Giving of Each State Each Year
data_total <- Giving %>%
  group_by(state) %>%
  summarise(giving2017 = sum(giving2017),
            giving2018 = sum(giving2018),
            giving2019 = sum(giving2019),
            giving2020 = sum(giving2020),
            giving2021 = sum(giving2021),
            .groups = "drop")

top_states <- data_total %>%
  pivot_longer(cols = starts_with("giving"),
               names_to = "year",
               values_to = "total_giving") %>%
  group_by(year) %>%
  top_n(5, total_giving) %>%
  arrange(year, desc(total_giving)) %>%
  ungroup()

# Select Colors
colors <- brewer.pal(12, "Paired")

# Create Dodged Bar Chart
p <- ggplot(top_states, aes(x = year, y = total_giving, fill = state)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top Donating States Each Year",
       x = "Year",
       y = "Total Giving") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = colors[1:10])

# Create Interactive Chart
ggplotly(p)
```

## Column {width="20%"}

### Row {height="40%"}

::: card
**Maps Insights**\
- California consistently led state giving each year from 2017 to 2020, peaking at 49.22% in 2017 but declining to 21.45% in 2021. This decline may reflect shifting donor priorities or external factors impacting donation behavior.\
- Tennessee emerged as a notable contributor in 2021, with a remarkable donation percentage of 15.53%. Factors such as successful advocacy campaigns or targeted fundraising efforts likely contributed to this trend.\
- Overall, regional dynamics played a significant role in state giving, with variations in donation percentages influenced by factors such as proximity to water-related issues, socioeconomic factors, and changing donor preferences. Adaptability in fundraising strategies and responsiveness to regional contexts emerged as key factors in maximizing charitable contributions.\
:::

```{r}
#| title: "State Giving Each Year"

# Grouping By State
create_us_map <- function(data, year) {
  map_data <- Giving %>%
    filter(!is.na(state)) %>%
    group_by(state) %>%
    summarise(total_giving = sum(get(paste0("giving", year)), na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(percentage_contribution = total_giving / sum(total_giving) * 100)
  
  # Creating Maps That Show Percentage of Giving
  plot_ly(
    data = map_data,
    type = 'choropleth',
    locationmode = 'USA-states',
    locations = ~state,
    z = ~percentage_contribution,
    text = ~paste("State: ", state, "<br>",
                  "Percentage of Total Giving: ", percentage_contribution, "%<br>"),
    colorscale = list(c(0, "rgb(202, 225, 255)"),   # Light blue
                      c(0.25, "rgb(107, 174, 214)"),
                      c(0.5, "rgb(49, 130, 189)"),
                      c(0.75, "rgb(8, 81, 156)"),
                      c(1, "rgb(8, 48, 107)")),     # Dark blue
    zmax = 50,  # Ensure color scale starts from 0
    zmin = 0,   # Set the maximum of color scale to 50
    marker = list(line = list(color = 'rgb(255,255,255)', width = 2)),
    colorbar = list(title = "Percentage of Total Giving")
  ) %>%
    layout(title = paste("Percentage Contribution of Each State to Total Giving in", year),
           geo = list(scope = 'usa'))
}
```

### Column {.tabset}

```{r}
#| title: "2017"
create_us_map(combined_data, 2017)
```

```{r}
#| title: "2018"
create_us_map(combined_data, 2018)
```

```{r}
#| title: "2019"
create_us_map(combined_data, 2019)
```

```{r}
#| title: "2020"
create_us_map(combined_data, 2020)
```

```{r}
#| title: "2021"
create_us_map(combined_data, 2021)
```

# Insights Page

## Row {height="20%"}

```{r}
#| include: false
combined_data <- merge(DS, Giving, by = "accountid", all = TRUE)
```

```{r}
#| content: valuebox
#| title: "DS 1-4 and DS 1-5 Lifetime Giving"

total_ds1_4_5_giving <- sum(combined_data$giving_lifetime[combined_data$ds_rating %in% c("DS1-4", "DS1-5")], na.rm = TRUE)

list(color = "green",
  value = dollar(total_ds1_4_5_giving)
)
```

```{r}
#| content: valuebox
#| title: "Very High Consistency Lifetime Giving"
combined_data <- combined_data %>%
  mutate(
    total_donations = rowSums(select(., starts_with("giving")) > 0),
    consecutive_donations = rowSums(select(., starts_with("giving")) > 0) == 5,
    consistency_rank = case_when(
      consecutive_donations ~ "Very High",
      total_donations >= 4 | total_donations == 3 ~ "High",
      total_donations == 2 | total_donations == 3 ~ "Medium",
      TRUE ~ "Low"
    )
  )
very_high_lifetime_giving <- sum(combined_data$giving_lifetime[combined_data$consistency_rank == "Very High"], na.rm = TRUE)

list(color = "yellow",
  value = dollar(very_high_lifetime_giving)
)
```

```{r}
#| content: valuebox
#| title: "California Lifetime Giving"

california_lifetime_giving <- sum(Giving$giving_lifetime[Giving$state == "CA"], na.rm = TRUE)

list(color = "blue",
  value = dollar(california_lifetime_giving)
)
```
